---
title: "Text Clustering (Custom Dataset)"
output:
  pdf_document: default
  html_notebook: default
---

## Reference

* https://rstudio-pubs-static.s3.amazonaws.com/445820_c6663e5a79874afdae826669a9499413.html
* http://mirlab.org/jang/books/dcpr/dcHierClustering.asp?title=3-2%20Hierarchical%20Clustering%20(%B6%A5%BCh%A6%A1%A4%C0%B8s%AAk)&language=chinese
* https://rpubs.com/skydome20/R-Note9-Clustering

## 0. Import Packages


```{r}

# Un-comment if you don't have these packages installed
# install.packages("tm")
# install.packages("proxy")
# install.packages("colorspace")

library(tm)
library(proxy)
library(colorspace)

```


## 1. Read Dataset

```{r}

df <- read.csv("product_description.csv", encoding = 'UTF-8', header = TRUE)
View(df)
```


```{r}
sentences <- df$description

tail(sentences)
```

## 2. Preprocessing the data

Create corpus object using text sentences:
```{r}
corpus = tm::Corpus(tm::VectorSource(sentences)) # corpus(=corpora): 語料庫, collections of documents text
```

Remove stopwords:
```{r}
print(tm::stopwords('english'))
corpus_cleaned <- tm::tm_map(corpus, tm::removeWords, tm::stopwords('english'))
```
Do Stemming for the words:
```{r}
corpus_cleaned <- tm::tm_map(corpus_cleaned, tm::stemDocument, language="english")
```

Trim excessive whitespaces:
```{r}
corpus_cleaned <- tm::tm_map(corpus_cleaned, tm::stripWhitespace)
```
View processed corpus:
```{r}
tail(corpus_cleaned$content)
```

## 3. Text Representation




Build Document-Term Matrix:
```{r}
doc_term_matrix <- tm::DocumentTermMatrix(corpus_cleaned)
doc_term_matrix
```

Calculate TF-IDF
```{r}
doc_term_matrix_tfidf <- tm::weightTfIdf(doc_term_matrix)
doc_term_matrix_tfidf
```

Remove sparse terms from a document-term matrix:
```{r}
doc_term_matrix_tfidf <- tm::removeSparseTerms(doc_term_matrix_tfidf, 0.999)
doc_term_matrix_tfidf
```

Load into native R matrix:
```{r}
tfidf_matrix <- as.matrix(doc_term_matrix_tfidf)
dim(tfidf_matrix)
```

Calculate distance(similarity) between documents using cosine similarity:
```{r}
# Cosine distance matrix (useful for specific clustering algorithms)
dist_matrix = proxy::dist(tfidf_matrix, method = "cosine")
dim(dist_matrix)
```

## 4. Run the clustering algorithms

```{r}
NUM_CLASS <- 3
```

1. K-means clustering:
```{r}
clustering_kmeans <- kmeans(dist_matrix, NUM_CLASS)
```

```{r}
clustering_kmeans
```

2. Hierarchical clustering:
```{r}
hierarchicals <- hclust(dist_matrix, method = "ward.D2")

hierarchical_clusters <- cutree(hierarchicals, k = NUM_CLASS)
```

```{r}
plot(hierarchicals)
```




## 5. Visualize clustered results



Use PCA to scale vectors' dimension (from document_n to 2):
```{r}
points <- cmdscale(dist_matrix, k = 2) # Running the PCA
dim(points)
```


```{r}
palette <- colorspace::diverge_hcl(NUM_CLASS) # Creating a color palette
previous_par <- par(mfrow=c(1,2), mar = rep(1, 4)) # partitionning the plot space
plot(points,
     main = 'K-Means clustering',
     col = as.factor(clustering_kmeans$cluster), xaxt = 'n', yaxt = 'n')
plot(points,
     main = 'Hierarchical clustering',
     col = as.factor(hierarchical_clusters), xaxt = 'n', yaxt = 'n')
```
